---
title: "Short Introduction to Spatial Data Processing in R"
author:
- "Richard Wen rrwen.dev@gmail.com"
date: "`r format(Sys.time(), '%B %d, %Y')`"
knit: |
    (function(input_rmd, ...) {
    rmarkdown::render(
        input_rmd,
        rmarkdown::html_document(
            toc = TRUE,
            toc_float = TRUE,
            highlight = "zenburn",
            code_folding = "show",
            df_print = "paged",
            self_contained = FALSE
        ),
        output_dir = "../docs",
        output_file = "index", ...)
    })
---

```{r include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
	warning = FALSE,
	message = FALSE
)
```

# Installation

1.  Install [R](https://www.r-project.org/)
2.  Install [RTools](https://cran.r-project.org/) if you are on Windows
3.  Install [RStudio](https://posit.co/download/rstudio-desktop/)

For more details, see [Software and Package Versions](#software-and-package-versions).

# Running This Code

1.  Ensure the installation steps above are completed
2.  Download a zip of the code and data [here](https://github.com/rrwen/r-spintro) repo```/archive/refs/heads/main.zip) and unzip it
    - Code Repository: [github.com/rrwen/r-sf-quickstart](https://github.com/rrwen/r-spintro)
3.  In RStudio, open the [src/src.Rproj](https://github.com/rrwen/r-spintro/blob/main/src/src.Rproj) file
4.  Then, open the [src/index.Rmd](https://github.com/rrwen/r-spintro/blob/main/src/index.Rmd) file
5.  In RStudio:
    - Run all code: Click the `Run` drop down (top right of the code pane) and click `Run All`
    - Generate HTML version: Click `knit` (top left of code pane) and a file will be generated in `docs/index.html`

# Libraries

Install R packages if needed:

* `tidyverse`: general data processing package that works with tabular structured data
* `lubridate`: convenient functions for handling date and time data
* `sf`: spatially enables the `tidyverse` package to extend functionality for spatial data
* `ggplot2`: package for creating plots, including maps of `sf` objects
* `tmap`: useful for quick static and interactive maps of `sf` objects
* `ggspatial`: makes plotting general map elements like base maps, north arrows and scale bars easier
* `prettymapr`: required dependency for `ggspatial` to create base maps

```{r eval = FALSE}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("sf")
install.packages("tmap")
install.packages("ggspatial")
install.packages("ggplot2")
install.packages("prettymapr")
```

```{r results = FALSE, echo = FALSE}

# Required packages
required_packages <- c(
  "ggspatial",
	"sf",
  "prettymapr",
  "tmap",
	"tidyverse",
  "lubridate",
  "ggplot2"
)

# Try to install packages if not installed
default_options <- options()
tryCatch(
	{
		# Disable interactivity
		options(install.packages.compile.from.source = "always")
		
		# Install package if not installed
		for (package in required_packages) {
			is_package_installed <- require(package, character.only = TRUE)
			if (!is_package_installed) {
				cat(paste0("Installing package: ", package, "\n"))
				install.packages(package)
			} else {
				cat(paste0("Package already installed: ", package, "\n"))
			}
		}
	},
	error = function(cond) {
		stop(cond)
	},
	finally = {
		options(default_options) # reset interactivity
	}
)
```

Load libraries.

```{r}
library(tidyverse)
library(lubridate)
library(sf)
library(ggspatial)
library(tmap)
library(ggplot2)
```

# Tutorial

For this tutorial, we will be using the following data mainly:

* [Toronto Police KSI Data](https://data.torontopolice.on.ca/datasets/TorontoPS::ksi/about)
* [Toronto Neighbourhoods Data](https://open.toronto.ca/dataset/neighbourhoods/)

Later in the tutorial, we will also use:

* [Toronto ASE Location Data](https://open.toronto.ca/dataset/automated-speed-enforcement-locations/)

## Reading Spatial Data

To read spatial data use the [read_sf](https://r-spatial.github.io/sf/reference/st_read.html) function.

**Note**: For shapefiles, simply use extension `.shp`.

```{r}
ksi <- read_sf("../data/toronto-ksi.geojson")
nbhoods <- read_sf("../data/toronto-nbhoods.geojson")
```

## Preview Non-spatial Data

Non-spatial data works similarly to standard `data.frame` structures, except the addition of a `geometry` column usually at the end of the columns.

Let's preview the non-spatial KSI data: 

```{r}
ksi
```

and also the Neighbourhoods data:

```{r}
nbhoods
```

Most [dplyr](https://dplyr.tidyverse.org/) functions from `tidyverse` will work with `sf` objects, allowing you to easily manipulate the non-spatial portions of the read spatial data.

For example, we get the number of KSI collisions each year using the `year` function from `lubridate` and `count` from `dplyr`:

```{r}
ksi %>%
  count(year(DATE))
```

## Preview Spatial Data

You can preview spatial data using [plot](https://r-spatial.github.io/sf/articles/sf5.html?q=plot#plot-methods-for-sf-and-sfc-objects), which generates several maps of up to the first 9 columns as the default.

Let's preview the KSI data:

```{r}
plot(ksi)
```

We can also preview a specific variable:

```{r}
plot(ksi["ACCLASS"])
```

and again for the Neighbourhoods data:

```{r}
nbhoods
```

## Preview Spatial Data Interactively

In addition to previewing spatial data statically, we can utilize the `tmap` package to preview spatial data interactively.

Let's try it with the KSI data (we limit it to the first 1000 records due to the size of the data):

* `tmap_mode` sets the visualization mode to interactive
* `tm_shape` tells the `tmap` package to use the `ksi` data
* `head` filters for the first 1000 rows
* `tm_dots` visualizes the data as point data

```{r}
tmap_mode("view") # set to interactive mode

tm_shape(ksi %>% head(1000)) +
  tm_dots(
    col = "ACCLASS", # color based on column
    clustering = TRUE, # group points together for cleaner visuals
    popup.vars = TRUE # allow user to click each point to inspect data
  )
```

We can also do the same for the Neighbourhoods data but with `tm_polygons` from `tmap` instead:

```{r}
tmap_mode("view")

tm_shape(nbhoods) +
  tm_polygons(
    	col = "#336699", # set polygon color
    	border.col = "white", # set border color
    	popup.vars = TRUE
    )
```

## Coordinate Reference Systems (CRS)

Each `sf` object should ideally have a defined CRS that affects the accuracy of coordinates, units of measurement, and spatially related calculations.

For the KSI data, we can view the CRS with [st_crs](https://r-spatial.github.io/sf/reference/st_crs.html), which is usually WGS84 (World Geodetic System 1984, a global standard that is often used in most spatial data) with public registry reference [EPSG:4326](https://epsg.io/4326):

```{r}
st_crs(ksi)
```

We also see the same CRS is used for the neighbourhoods data:

```{r}
st_crs(nbhoods)
```

In most situations, you do not need to modify or define the CRS as the `sf` package handles accurate calculations and most CRS are distributed in WGS84, but we should always double check.

To define the CRS if it is not given in the spatial data, we use [st_set_crs](https://r-spatial.github.io/sf/reference/st_crs.html) with the correct EPSG number (`4326` in our case), which can be found on [espg.io](https://epsg.io).

As an example, we will use a separate example such as the `nc.shp` file from the `sf` package, but set the CRS to `NA` as a demonstration:

```{r}

# Read example data
exdata <- read_sf(
  system.file("shape/nc.shp", package="sf"),
  crs = NA
)

# View CRS
st_crs(exdata)
```

Now can set the CRS to the correct one, which is NAD27 with [EPSG:4267](https://epsg.io/4267):

```{r}

# Set CRS to NAD27
exdata <- exdata %>%
  st_set_crs(4267)

# View CRS again after setting it to NAD27
st_crs(exdata)
```

Now that there is a defined CRS, we can reproject them with [st_transform](https://r-spatial.github.io/sf/reference/st_transform.html) to a different CRS, such as WGS84 (EPSG:4326), in case we do not have consistent CRS across different spatial data:

```{r}
exdata %>%
  st_transform(4326)
```

It is important to ensure that all your spatial data is in the same CRS before mapping or conducting any spatially related calculations or processes.

## Spatial Buffers

A common spatial data processing task is to apply buffers (radial extensions of spatial geometries) to spatial objects, which can be done using the [st_buffer](https://r-spatial.github.io/sf/reference/geos_unary.html) function.

Let's try this for a couple of records in the KSI data, which are originally points as seen below:

```{r}

# Select the first 3 records of KSI
ksi_buff <- ksi %>%
  head(3)

# Display the KSI geometry as points
ksi_buff %>%
  select(geometry) %>%
  plot
```

Then we can buffer the points in the KSI data by 1000 meters (unit of measure follows the CRS, which is meters for WGS84):

```{r}

# Buffer the point geoms by 1000m
ksi_buff <- ksi_buff %>%
  select(geometry) %>%
  st_buffer(dist = 1000)

# Plot polygon geometries after buffer
ksi_buff %>%
  select(geometry) %>%
  plot
```

Applying buffers to points is useful for capturing other spatial objects within a certain distance of each point - relating captured spatial objects to each point.

## Spatial Joins

Another commonly used spatial function is spatial joins, which can be done using the [st_join](https://r-spatial.github.io/sf/reference/st_join.html) function.

Let's try getting the number of KSI collisions in 2022 within each Neighbourhood.

First, we filter out all KSI collisions for year 2022:

```{r}

# Filter for 2022 data
ksi_2022 <- ksi %>%
  filter(year(DATE) == 2022)

# Preview the data
ksi_2022 %>% as_tibble
```

Then, we must join each KSI point to each neighbourhood polygon:

```{r}

# Join KSI data to be within each neighbourhood polygon
nbhoods_join <- nbhoods %>%
  st_join(ksi_2022)

# Preview the joined data
nbhoods_join %>% as_tibble
```

Finally, we can apply the `count` function to get the number of collisions per neighbourhood in column `n`:

```{r}

# Count the joined ksis per nbhood
nbhoods_ksi_2022 <- nbhoods_join %>%
  count(AREA_ID, AREA_NAME)

# Preview ksi per nbhood in 2022
nbhoods_ksi_2022 %>% as_tibble
```

We can also plot these spatially:

```{r}
plot(nbhoods_ksi_2022["n"])
```

## Making Maps for Publications

The `plot` function is convenient for previewing spatial data quickly, but makes it difficult to add standard map elements such as base maps, north arrows and scale bars.

For publications, we can use the `ggspatial` package to create publication ready maps.

As an example, we will utilize our joined 2022 KSI collision counts for each neighbourhood in Toronto, and create a publication ready map.

We will also load the Automated Speed Enforcement (ASE) data for Toronto to map an extra layer on top of the neighbourhood collision data.

```{r}
ase <- read_sf("../data/toronto-ase.geojson")
```

Now that we have all the data we need, let's create a publication ready map:

* [annotation_map_tile](https://paleolimbot.github.io/ggspatial/reference/annotation_map_tile.html): creates a base map
* [annotation_north_arrow](https://paleolimbot.github.io/ggspatial/reference/annotation_north_arrow.html): creates the north arrow
* [annotation_scale](https://paleolimbot.github.io/ggspatial/reference/annotation_scale.html): creates the scale bar
* [layer_spatial](https://paleolimbot.github.io/ggspatial/reference/layer_spatial.html): creates a spatial layer
* [labs](https://ggplot2.tidyverse.org/reference/labs.html): controls plot label elements
* [guides](https://ggplot2.tidyverse.org/reference/guides.html): controls legend elements
* [theme](https://ggplot2.tidyverse.org/reference/ggtheme.html): controls overall map/plot elements

```{r}
ggplot() +
  annotation_map_tile(
    zoomin = 1, # affects resolution of base map tiles
    type = "cartolight", # type of base map
    cachedir = "../data/cache" # cache the map tiles to avoid redownload
  ) +
  annotation_north_arrow(
    width = unit(0.2, "cm"),
    height = unit(0.5, "cm"),
    location = "tr" # where to place the arrow
  ) +
  annotation_scale(
    style = "ticks",
    location = "br" # where to place the scale bar
  ) +
  labs(
    title = "2022 Neighbourhood Collisions (Toronto, ON, CA)", # map title
    caption = "Data Source(s): Toronto Police Service Public Safety Portal, City of Toronto Open Data\nCRS: WGS84 / *Automated Speed Enforcement (ASE), Killed or Seriously Injured (KSI)" # map note
  ) +
  layer_spatial( # add spatial data layer for collisions
    nbhoods_ksi_2022, 
    aes(fill = n) # set the fill color to n, which is num of collisions
  ) +
  layer_spatial( # add spatial data layer for ASE
    ase,
    fill = NA,
    aes(color = "ASE* Camera") # set name of points for ASE
  ) +
  scale_color_manual(
    values = c("ASE* Camera" = "salmon") # change ase point color
  ) +
  guides(
    fill = guide_legend(
      order = 1,
      title = "KSI* Collisions" # legend title for collisions
    ),
    color = guide_legend(
      order = 2,
      title = element_blank() # remove legend title for ase
    )
  ) +
  theme(
    plot.caption = element_text(size = 6, color = "darkgray"), # alter note text
    legend.key = element_blank(), # remove grey legend background in ase points
    plot.title = element_text(hjust = 0.5) # center map title
  )
```

The visualization element of spatial data is relatively more complicated, thus other alternatives to creating maps more interactively are:

* [QGIS](https://www.qgis.org/en/site/): free and open source GIS software available for all operating systems
* [ArcGIS Pro](https://www.esri.com/en-us/arcgis/products/arcgis-pro/overview): paid industry GIS software for Windows

## Writing Spatial Data

Now that we are done processing our data, we should save the results to a file with the [write_sf](https://r-spatial.github.io/sf/reference/st_write.html) function.

Let's save our joined 2022 KSI collision counts for each neighbourhood in the `data` folder:

```{r}

# Save a geojson file
nbhoods_ksi_2022 %>%
  write_sf("../data/toronto-nbhoods-ksi-2022.geojson")

# Save a shapefile
nbhoods_ksi_2022 %>%
  write_sf("../data/toronto-nbhoods-ksi-2022.shp")
```

Later on, we can read the saved shapefile back in:

```{r}

# Read the shapefile back in
nbhoods_ksi_2022 <- read_sf("../data/toronto-nbhoods-ksi-2022.shp")

# Preview the data
nbhoods_ksi_2022 %>% as_tibble
```

# Other Tutorials

That concludes the tutorial! For further information and other tutorials, see:

* [sf: Simple Features for R](https://r-spatial.github.io/sf/articles/sf1.html)
* [ggspatial: Spatial objects for ggplot2](https://paleolimbot.github.io/ggspatial/articles/ggspatial.html)
* [ggplot2: Introduction to ggplot2](https://ggplot2.tidyverse.org/articles/ggplot2.html)
* [dplyr: Introduction to dplyr](https://dplyr.tidyverse.org/articles/dplyr.html)

# Software and Package Versions {#software-and-package-versions}

R and R package versions:

```{r echo = FALSE}
sessionInfo()
```

RStudio version:

```{r echo = FALSE}
rstudioapi::versionInfo()
```
