---
title: "Quickstart to Mapping using the sf R Package"
author:
- "Richard Wen richard.wen@utoronto.ca"
date: "`r format(Sys.time(), '%B %d, %Y')`"
knit: |
    (function(input_rmd, ...) {
    rmarkdown::render(
        input_rmd,
        rmarkdown::html_document(
            toc = TRUE,
            toc_float = TRUE,
            highlight = "zenburn",
            code_folding = "hide",
            df_print = "paged",
            self_contained = FALSE
        ),
        output_dir = "../docs",
        output_file = "index", ...)
    })
---

```{r include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
	warning = FALSE,
	message = FALSE
)
```

# Installation

1.  Install [R](https://www.r-project.org/)
2.  Install [RTools](https://cran.r-project.org/) if you are on Windows
3.  Install [RStudio](https://posit.co/download/rstudio-desktop/)

For more details, see [Software and Package Versions](#software-and-package-versions).

# Running This Code

1.  Ensure the installation steps above are completed
2.  Download a zip of the code and data [here](https://github.com/rrwen/r-sf-quickstart) repo```/archive/refs/heads/main.zip) and unzip it
    - Code Repository: [github.com/rrwen/r-sf-quickstart](https://github.com/rrwen/r-sf-quickstart)
3.  In RStudio, open the [src/src.Rproj](https://github.com/rrwen/r-sf-quickstart/blob/main/src/src.Rproj) file
4.  Then, open the [src/index.Rmd](https://github.com/rrwen/r-sf-quickstart/blob/main/src/index.Rmd) file
5.  In RStudio:
    - Run all code: Click the `Run` drop down (top right of the code pane) and click `Run All`
    - Generate HTML version: Click `knit` (top left of code pane) and a file will be generated in `docs/index.html`

# Libraries

Install R packages if needed.

```{r eval = FALSE}
install.packages("tidyverse")
install.packages("sf")
install.packages("ggspatial")
```

```{r results = FALSE, echo = FALSE}

# Required packages
required_packages <- c(
  "ggspatial",
	"sf",
	"tidyverse"
)

# Try to install packages if not installed
default_options <- options()
tryCatch(
	{
		# Disable interactivity
		options(install.packages.compile.from.source = "always")
		
		# Install package if not installed
		for (package in required_packages) {
			is_package_installed <- require(package, character.only = TRUE)
			if (!is_package_installed & package != "osmplotr") {
				cat(paste0("Installing package: ", package, "\n"))
				install.packages(package)
			} else {
				cat(paste0("Package already installed: ", package, "\n"))
			}
		}
	},
	error = function(cond) {
		stop(cond)
	},
	finally = {
		options(default_options) # reset interactivity
	}
)
```


# Software and Package Versions {#software-and-package-versions}

R and R package versions:

```{r echo = FALSE}
sessionInfo()
```

RStudio version:

```{r echo = FALSE}
rstudioapi::versionInfo()
```
